---
title: "Spatial Unit Root Diagnostics with spatialunitroot"
author: "Bas Machielsen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Unit Root Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(spatialunitroot)
library(fixest)
```

## Introduction

This vignette demonstrates the `spatialunitroot` R package, which implements the spatial unit root diagnostic tests and transformations of Müller & Watson (2024, *Econometrica*). The package is an R port of the SPUR Stata package by Becker, Boll & Voth (2025, *Stata Journal*).

Spatial data often exhibit strong dependence across locations. Just as time series with unit roots can produce spurious regressions, spatial processes with (near) unit roots can lead to misleading inference. This package provides tools to:

1. **Diagnose** spatial unit roots using I(1) and I(0) tests
2. **Remove** spatial unit roots via spatial differencing transformations
3. **Characterize** spatial persistence via half-life confidence intervals

Throughout, we replicate the numbers published in **Appendix A (Table A.1) of Becker, Boll & Voth (2025)**, which reproduce the Chetty et al. (2014) results of Müller & Watson (2024). The dataset used there ships with this package, so the published table can be reproduced directly.

## The Chetty et al. (2014) Commuting Zone Data

We use the dataset from Chetty, Hendren, Kline & Saez (2014), which contains measures of intergenerational mobility and its correlates across 722 US commuting zones. This is the same data the Stata package uses in its Appendix A.

```{r data}
data(chetty)
str(chetty[, 1:10], vec.len = 2)
```

Spatial coordinates are stored in `s_1` (latitude) and `s_2` (longitude), following the convention of Müller & Watson's `scpc` package. When `latlong = TRUE`, distances are computed as great-circle distances on the sphere.

## Diagnosing Spatial Unit Roots

We illustrate the procedure using absolute upward mobility (`am`), the key outcome variable in Chetty et al. (2014) and the first row of Table A.1.

### I(1) Test: Is there a spatial unit root?

The I(1) test has the null hypothesis of a spatial unit root. A high p-value means we cannot reject the presence of a unit root.

```{r i1-test}
set.seed(42)
t_i1 <- spurtest(am ~ 1, data = chetty, coords = ~ s_1 + s_2,
                 type = "i1", latlong = TRUE, q = 15, nrep = 10000)
summary(t_i1)
```

With a p-value of `r round(t_i1$p_value, 2)`, we **fail to reject** the spatial unit root null. Table A.1 of Becker, Boll & Voth (2025) reports an I(1) p-value of **0.38** for absolute mobility — the small difference is Monte Carlo simulation noise.

### I(0) Test: Is the variable spatially stationary?

The I(0) test has the null hypothesis of **no** spatial unit root (spatial stationarity). A low p-value indicates strong evidence *against* spatial stationarity.

```{r i0-test}
set.seed(42)
t_i0 <- spurtest(am ~ 1, data = chetty, coords = ~ s_1 + s_2,
                 type = "i0", latlong = TRUE, q = 15, nrep = 10000)
summary(t_i0)
```

The p-value is essentially zero, so we **strongly reject** the null that `am` is I(0). This matches Table A.1, which reports an I(0) p-value of **0.00** for absolute mobility. Together, the I(0) and I(1) tests provide strong evidence that `am` has a spatial unit root.

### Spatial Half-Life

The half-life is the distance at which spatial correlation drops to one-half. We report it in normalized units (fractions of the maximum pairwise distance), as in Table A.1.

```{r halflife}
set.seed(42)
hl <- spurhalflife(am ~ 1, data = chetty, coords = ~ s_1 + s_2,
                   latlong = TRUE, normdist = TRUE, q = 15, nrep = 10000)
summary(hl)
```

The half-life confidence interval for `am` is approximately [`r round(hl$ci_lower, 2)`, ∞] in fractions of the maximum pairwise distance, matching the published interval of **[0.09, ∞]**. The unbounded upper bound is a hallmark of a (near) unit root.

## Reproducing Table A.1

We now assess spatial persistence across a representative set of the variables in Table A.1, comparing the R output to the published Stata results. For each variable we report the I(1) and I(0) test p-values and the half-life confidence interval, with the value from Becker, Boll & Voth (2025) in parentheses.

```{r multi-test, results="asis"}
# Variable -> published Table A.1 values: c(I1, I0, HL_lower, HL_upper)
paper <- list(
  am        = c(0.38, 0.00, 0.09, Inf),
  fracblack = c(0.11, 0.01, 0.04, Inf),
  racseg    = c(0.01, 0.13, 0.00, 0.28),
  segpov25  = c(0.28, 0.03, 0.05, Inf),
  fraccom15 = c(0.57, 0.00, 0.14, Inf),
  hipc      = c(0.13, 0.14, 0.02, Inf),
  gini      = c(0.78, 0.00, 0.26, Inf),
  tsr       = c(0.23, 0.13, 0.05, Inf),
  hsdrop    = c(0.09, 0.02, 0.03, Inf),
  scind     = c(0.72, 0.00, 0.22, Inf),
  manshare  = c(0.20, 0.00, 0.06, Inf),
  tlfpr     = c(0.51, 0.00, 0.12, Inf),
  fracfor   = c(0.56, 0.04, 0.17, Inf)
)

fmt_inf <- function(x) if (is.infinite(x)) "∞" else sprintf("%.2f", x)

results <- do.call(rbind, lapply(names(paper), function(v) {
  set.seed(42)
  t1 <- spurtest(as.formula(paste(v, "~ 1")), data = chetty,
                 coords = ~ s_1 + s_2, type = "i1",
                 latlong = TRUE, q = 15, nrep = 5000)
  set.seed(42)
  t0 <- spurtest(as.formula(paste(v, "~ 1")), data = chetty,
                 coords = ~ s_1 + s_2, type = "i0",
                 latlong = TRUE, q = 15, nrep = 5000)
  set.seed(42)
  hl <- spurhalflife(as.formula(paste(v, "~ 1")), data = chetty,
                     coords = ~ s_1 + s_2, latlong = TRUE,
                     normdist = TRUE, q = 15, nrep = 5000)
  p <- paper[[v]]
  data.frame(
    Variable    = v,
    `I(1) p`    = sprintf("%.2f (%.2f)", t1$p_value, p[1]),
    `I(0) p`    = sprintf("%.2f (%.2f)", t0$p_value, p[2]),
    `HL lower`  = sprintf("%.2f (%.2f)", hl$ci_lower, p[3]),
    `HL upper`  = sprintf("%s (%s)", fmt_inf(hl$ci_upper), fmt_inf(p[4])),
    check.names = FALSE
  )
}))

knitr::kable(
  results,
  caption = "Reproducing Table A.1 of Becker, Boll & Voth (2025). Each cell shows the R result with the published Stata value in parentheses."
)
```

The R implementation reproduces the published p-values and half-life intervals to within Monte Carlo error. As Table A.1 shows, the overwhelming majority of the Chetty et al. (2014) variables fail to reject the I(1) null and reject the I(0) null, indicating strong spatial persistence.

## Correcting Spurious Regression

A key application is diagnosing and correcting spurious spatial regression. Following the algorithm in Figure 5 of Becker, Boll & Voth (2025): once a variable cannot be rejected as I(1), we difference *all* variables (dependent and independent) before re-estimating. We illustrate the workflow with a regression of `am` on `fracblack` and `racseg`.

### Step 1: Naïve OLS

```{r ols-naive}
m_ols <- feols(am ~ fracblack + racseg, data = chetty, se = "standard")
summary(m_ols)
```

### Step 2: Test residuals for a spatial unit root

```{r resid-test}
set.seed(42)
t_resid <- spurtest(m_ols, coords = ~ s_1 + s_2,
                    type = "i1resid", latlong = TRUE,
                    q = 15, nrep = 5000)
t_resid
```

The residual test fails to reject a spatial unit root in the residuals, so the OLS inference above is not reliable and we proceed to difference all variables.

### Step 3: Apply the LBM-GLS transformation

```{r transform}
transformed <- spurtransform(~ am + fracblack + racseg, data = chetty,
                              coords = ~ s_1 + s_2,
                              prefix = "h_", latlong = TRUE)
```

### Step 4: Re-estimate on transformed data

```{r ols-transformed}
m_trans <- feols(h_am ~ h_fracblack + h_racseg, data = transformed,
                 se = "standard")
summary(m_trans)
```

The R² drops from `r round(fixest::r2(m_ols)["r2"], 3)` to `r round(fixest::r2(m_trans)["r2"], 3)` after removing spatial trends, indicating that much of the apparent relationship was driven by shared spatial patterns — a classic spurious regression result. In practice, one would additionally apply SCPC inference (Müller & Watson 2022, 2023) to the transformed regression to account for any remaining weak spatial correlation.

## Comparison with the Stata SPUR Package

The R implementation reproduces the Stata SPUR package (Becker, Boll & Voth, 2025). The table above replicates Table A.1 across variables; the key checks for absolute mobility (`am`) are:

| Statistic | Stata (Table A.1) | R (`spatialunitroot`) |
|-----------|-------------------|------------------------|
| I(1) test p-value | 0.38 | `r round(t_i1$p_value, 2)` |
| I(0) test p-value | 0.00 | `r round(t_i0$p_value, 2)` |
| Half-life 95% CI (norm.) | [0.09, ∞] | [`r round(hl$ci_lower, 2)`, ∞] |

Minor numerical differences arise from Monte Carlo simulation noise (of order $1/\sqrt{n_{rep}}$) and from differences between the R and Stata random number generators, but the qualitative conclusions — and the reported values to two decimal places — are identical.

## References

- Becker, S. O., Boll, P. D., & Voth, H.-J. (2025). "Testing and Correcting for Spatial Unit Roots in Regression Analysis." *Stata Journal*.
- Chetty, R., Hendren, N., Kline, P., & Saez, E. (2014). "Where is the Land of Opportunity?" *Quarterly Journal of Economics*, 129(4), 1553–1623.
- Müller, U. K. & Watson, M. W. (2024). "Spatial Unit Roots and Spurious Regression." *Econometrica*, 92, 1661–1695.
